---
title: "Assignment 1"
output:
  pdf_document: default
  html_notebook: default
---

```{r include=FALSE}
library(MASS)
library(tinytex)
library(rms)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(aod)
```


```{r}
ars_data <- read.table("arsenic-2.txt", header = TRUE)
head(ars_data)
```
a) Producing plots



```{r}
#plot(ars_data$assoc, ars_data$dist, ylab = "Dist (in meters) to safest well", xlab = "Active community members")
boxplot( dist~assoc,
         data=ars_data,
         main="Distance to safe well vs. Active community members ",
         xlab="Active community members (assoc)",
         ylab="Distance to the closest known safe well (meters)",
         col="blue",
         border="black"
       )
```

We notice when fitting distance to the closest known safe well and whether 
members of the household are active community members, that there isn't much
difference. Citizens who aren't active community members travel more of a 
distance compared to those who are but not by a significant margin. 
Both groups in general seem to travel around the same distance indicated by the
median. 
The only concern is the outlier existing in the group of citizens that aren't 
active community members. This individual traveled almost 350 meters.



b) 

Logistic regression model with all additive predictors of arsenic, distance
traveled in meters, 
whether members of the household were an active community member and education. 
```{r}
ars_predicts <- glm(formula = switch ~ arsenic + dist + assoc + educ , 
                    family = binomial(link = "logit"), data = ars_data)
ars_predicts
```



Logistic regression model with all additive predictors and an 
interaction between distance traveled in meters and whether members of the 
house hold was an active community member.
```{r}

ars_interact <- glm(formula = switch ~ arsenic + dist + assoc + educ 
                    + dist*assoc, 
                    family = binomial(link = "logit"), data = ars_data)
ars_interact
```

We computer the brier score for the original data for both models
```{r}
predicted_predicts.model <- predict(ars_predicts, ars_data, type = "response")
brierscore_predicts.model <- (mean((predicted_predicts.model - ars_predicts$residuals)^2))/nrow(ars_data)
brierscore_predicts.model
```

```{r}
predicted_interact.model <- predict(ars_interact, ars_data, type = "response")
brierscore_interact.model <- (mean((predicted_interact.model - ars_interact$residuals)^2))/nrow(ars_data)
brierscore_interact.model
```

The model with all additive predictors no interactions has the better Brier 
score but not by that much. The additive predictors model has a lower Brier 
score of 0.01072395 in comparison to 0.01073027 of the interaction model 
of distance and 
active community member


#Question-C. 

We generate a random sample that has the same number of rows from our orginal
data which is about 500.
which
essientially gives us the index.train data from the validate function.
```{r}
set.seed(1002625448)
random_sample <- sample(x=1:nrow(ars_data), size = nrow(ars_data), replace = TRUE)
arsenic_sample <- ars_data[random_sample,]
```

We will now refit a model with all additive predictors and a model with
interactions agian using this sample. 

```{r}
ars_random.predicts <- glm(formula = switch ~ arsenic + dist + assoc + educ , 
                    family = binomial(link = "logit"), data = arsenic_sample)
ars_random.predicts

```

Same thing fro the interaction model
```{r}
ars_random.interact <- glm(formula = switch ~ arsenic + dist + assoc + educ 
                    + dist*assoc, 
                    family = binomial(link = "logit"), data = arsenic_sample)
ars_random.interact
```

Brier Score computations
```{r}
predicted_predicts.ran <- predict(ars_random.predicts, arsenic_sample, type = "response")
brierscore_predicts.new.model <- (mean((predicted_predicts.ran - ars_random.predicts$residuals)^2))/nrow(arsenic_sample)
brierscore_predicts.new.model
```

```{r}
predicted_interact.ran <- predict(ars_random.interact, arsenic_sample, type = "response")
brierscore_interact.new.model <- (mean((predicted_interact.ran - ars_random.interact$residuals)^2))/nrow(arsenic_sample)
brierscore_interact.new.model
```

#Question-D.
We now run the same computations of this sample data on the original data, which
essientially gives us the index.text data from the validate function.
```{r}
predicted_predicts.ran2 <- predict(ars_random.predicts, ars_data, type = "response")
brierscore_predicts.new.model2 <- (mean((predicted_predicts.ran2 - ars_random.predicts$residuals)^2))/nrow(ars_data)
brierscore_predicts.new.model2
```

```{r}
predicted_interact.ran2 <- predict(ars_random.interact, ars_data, type = "response")
brierscore_interact.new.model2 <- (mean((predicted_interact.ran2 - ars_random.interact$residuals)^2))/nrow(ars_data)
brierscore_interact.new.model2
```

Overall, we find that the model with additive predictors always produces
the better brier score than the interaction model. Thus, the additive predictors
model is the best model.


#Question 2
a) 

loading and looking at data
```{r}
hiv_data <- read.table("hiv-1.txt", header = TRUE)
head(hiv_data)
```

After viewing we convert sex from integer to a factor 
```{r}
hiv_data$sex <- as.factor(hiv_data$sex)
unique(hiv_data$sex)
```

We also convert intervention into a factor format as well 
```{r}
hiv_data$intervention <- as.factor(hiv_data$intervention)
unique(hiv_data$intervention)
```

We want to do some exploratory analysis here by getting a better understanding
of our outcome variable; fupacts. Fupacts is the number of unprotected sexual
acts after the intervention
```{r}
hiv_data %>% ggplot(aes(x=fupacts))+geom_histogram()
```

Based off this histogram we notice we're going to need a discrete distribution
where values are defined at positive values.
The poisson distribution fits this criteria

```{r}
boxplot( fupacts ~ intervention ,
         data=hiv_data,
         main="Intervention effectiveness on number of unprotected sexual acts",
         xlab="Intervention Type",
         ylab="# of unprotected sexual acts after intervention",
         col="blue",
         border="black"
       )
```

Exploring this data with number of unprotected sexual acts after the 
intervention vs the intervention type seems to tell us that having both partners
counselled and having the woman counselled reduces the number of unprotected 
sexual acts.
Particularly, having the woman counselled produces the fewest amount as we can 
see by the median line on the boxplot. 

```{r}
hiv_data_clean <- hiv_data %>% slice(-c(140))
hiv_data_clean
```


Noticed the outlier on the women counselled boxplot where about 200 sexual acts
were committed seems unreasonable so we will remove this.










b) 

As previously stated, we're going to need a discrete distribution
where values are defined at positive values since this is a count outcome.
Also, remember our outcome variable fupacts is an unrestricted count. 




c) Fit a Poisson regression model

```{r}
hiv_poireg <- glm(data = hiv_data_clean, formula = fupacts ~ sex + bs_hiv 
                  + intervention, 
                  offset = log(bupacts), family = poisson)
summary(hiv_poireg)
```

We fit a poisson model with number of unprotected sexual acts after the 
intervention (fupacts) as the outcome variable. Sex, HIV status, 
and intervention as predictors.
We use the number of unprotected sexual acts before the intervention (bupacts)
as the offset as we want to know the number of unprotected sexuals acts after
the intervention in proportion to the number of how much they were doing before.
We use log to make sure R doesn't complain. 
We notice an interaction between having hiv and a woman being counselled

```{r}
summary(hiv_poireg)
```








d) 

We notice a cone/fanning out shape when graphically displays the residuals,
which is problematic. This indicates evidence of overdispersion.

```{r}
hiv_stres <- rstandard(hiv_poireg)
test_stat <- sum(hiv_stres^2)

plot(x= hiv_poireg$fitted.values, y= hiv_stres,
xlab = "Fitted Values",
ylab = "Standardized Residuals",
main = "Standardized Residuals vs. Fitted Values")
```

Numerically assessing for overdispersion confirms this. 
```{r}
degrees_freedom_hiv = nrow(hiv_data_clean) - length(hiv_poireg$coefficients)

1-pchisq(test_stat, degrees_freedom_hiv)
```
Our test_stat is so low that it renders as 0. This indicates evidence
of overdispersio as it's  below a p-value of 0.05


e) 

We run a negative binomial model to account for this overdispersion as this has
a different form than the poisson. 
We also notice our estimates are close as our possion model but not the same.
```{r}
nb_model <- glm.nb(fupacts ~ sex + bs_hiv + intervention  
                    + offset(log(bupacts)), data = hiv_data_clean)

hiv_nbstres <- rstandard(nb_model)
test_stat <- sum(hiv_nbstres^2)

plot(x= nb_model$fitted.values, y= hiv_nbstres,
xlab = "Fitted Values",
ylab = "Standardized Residuals",
main = "Standardized Residuals vs. Fitted Values")
```

```{r}
summary(nb_model)
```


f) 

It seems as if have both counselled proves to be the most effective as 
no counselling gives a p-value of 0.24 and woman counselled gives a p-value of
0.71424.

